---
title: "index"
author: "Jan Sodoge"
date: "1/2/2021"
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



## aweSOM



__aweSOM__ is an R package offering a set of tools to explore and analyze datasets with Self-Organizing Maps (also known as [Kohonen maps](https://en.wikipedia.org/wiki/Self-organizing_map)), a form of artificial neural network originally created by [Teuvo Kohonen](https://en.wikipedia.org/wiki/Teuvo_Kohonen) in the 1980s. The package introduces interactive plots, making analysis of the SOM easier.

__aweSOM__ provides a variety of functions to analyze and visualize SOMs. They can be used either through the web-based interface (called by ```aweSOM()```) or through command-line functions. The package relies on the ```kohonen``` package for the training of SOMs.



### Install

Install the developer version of aweSOM from Github using the devtools package

```{r, eval = FALSE}
devtools::install_github("jansodoge/awesom_dev_version")
```


### aweSOM web-based interface

An intuitive and user-friendly approach to training and analyzing datasets with self-organizing maps is enabled by the aweSOM web-based interface. It allows the user to import data, train a SOM, analyze via the interactive visualizations, download results, and access R-code to replicate results. The web-based interface can be launched using the following function.

```{r, eval = FALSE}
aweSOM::aweSOM()
```



### aweSOM command-line functions

The aweSOM package provides command-line functions that enable functionality similar to the web-based interface. Detailed documentation of the command-line functions and their usage within a typical workflow is documented within the vignette, [here](articles/aweSOM-package.html)